Estimating the Initial Susceptible Fraction in Dengue Dynamics

### 2014-11-06 draft_lm1.tex
finished rewriting the introduction

The attack ratio is also influenced by the reproductive number of the disease 
\cite{bacaer_final_2009, katriel_attack_2012}. With Dengue, since transmission 
is affected by environmental conditions as discussed above, the reproductive 
number is a function of time.

@@

The initial number of susceptibles -- $S_0$ -- is a a very powerful parameter 
with regards to epidemic risk.
Its value affects directly the local $R_0$, thus being intrumental in predicting 
epidemics, but it also affects the speed and size of the epidemic.

In this paper, we estimate the number of susceptibles, allowing for 
discontinuities right before every epidemic in the last 18 years. 
Through this method we hope to detect drastic changes in the mixture of 
susceptibles cause by the replacement of the predominant serotype, which 
can redefine susceptibility for the following season.

To accomplish this we use a simplified model of 
dengue transmission based on a Susceptible-infecctious-Removed (SIR) model. 
Since Dengue is not caused by a single pathogen, but by four separate strains 
with limited cross-immunity, when we talk about number of susceptibles, in the 
context of a single serotype SIR model, we are actually referring to the sum of 
the susceptibles to the currently circulating serotypes at any given time.

Some information is available about the predominant serotypes during the 
period of study\cite{macedo_virological_2013}.

The estimation of the  number of susceptibles have been attempted 
before, for other diseases\cite{bjornstad_dynamics_2002, 
wallinga_reconstruction_2003}. These methods try to reconstruct the entire 
series of infectious and susceptibles for measles 
outbreaks from case data.\todo[inline]{anyone else?} In the case of Dengue the 
series of susceptibles to all possible serotypes, cannot be reconstructed based 
solely on a deterministic transmission model, since the arrival/re-emergenge of 
new serotypes (which are a stochastic events) can change drastically the pool 
of susceptibles throwing off any sequential estimation of the dynamics. 
Adopting a multi-strain transmission model, does not help because of 
insufficient information about the proportions of each serotype in case data.

>>>>

The attack ratio is also influenced by the reproductive number of the disease 
\cite{bacaer_final_2009, katriel_attack_2012}. With Dengue, since transmission 
is affected by environmental conditions as discussed above, the reproductive 
number is also expected to vary with time.

@@

In this paper, we estimate the number of susceptibles before every major 
outbreak ($S_{0, year}$) in the last 18 years. Due to the alternation of 
ciculating serotypes, we detected fluctuating levels of susceptibility the 
circulating virus(es). To Estimate the susceptibles we have we have fixed the 
tramissibility based on the $R_t$ series derived from data\cite{nishiura}. Then, 
from the incidence series and the population at risk, we could then calculate 
the attack ratio for each epidemic.

To accomplish the estimation of the $S_0$, we use a simplified model of 
dengue transmission based on a single-strain Susceptible-infecctious-Removed 
(SIR) model. The reason for the choice of a single strain model, is the lack of 
accurate information regarding the actual proportions of each virus in 
circulation. However, some information is available about the predominant 
serotypes for some epidemics in the period 
of study\cite{macedo_virological_2013}.

The estimation of the  number of susceptibles have been attempted 
before, for other diseases\cite{bjornstad_dynamics_2002, 
wallinga_reconstruction_2003}. These methods try to reconstruct the entire 
series of infectious and susceptibles for measles 
outbreaks from case data.\todo[inline]{anyone else?} In the case of Dengue the 
series of susceptibles to all possible serotypes, cannot be reconstructed based 
solely on a deterministic transmission model, since the arrival/re-emergenge of 
new serotypes (which are a stochastic events) can change drastically the pool 
of susceptibles throwing off any sequential estimation based on the incidence
dynamics.
